Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-06-10 paper/main.tex
top rework

\section{Related Work}
We build on the exentsive work codifying the structure of visualization data to develop a more generalizable method for expressing the structure that visualizations are expected to preserve. Broadly, previous work describes structure as a combination of the topological properties of the data, the mathematical structure of the data fields, and equivalent changes to data and graphics. We propose that a unified abstraction model can guide developers in building structure aware composable, reusable software library components with more consistent interfaces.

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{
    A topological representation of the data (in purple) is a simplified representation of how the records in a dataset are connected. The table is a set of independent records; therefore it has a 0D continuity represented by the disconnected dots. These records can also be samples from a timeseries, which has a 1D continuity. \note{started making a map of the temperature data but was taking a while so I didn't  }
    \label{fig:related-work:continuity:ktypes}}
\end{figure}

Roughly, topology is a formal way of encoding the connectivity of the elements of the dataset \cite{wilkinsonGrammarGraphics2005} and there is an expectation that the visual elements should match this connectivity. Formally, a topological space is a set $X$ with a structure called a topology. The set $X$ is a collection of points, where these points are \note{getting tripped up on how much of the definition to include here, but this is the natural place to give a definition of topology and opensets}
Some common topologies, as illustrated in\autoref{fig:related-work:continuity:ktypes}, are point, lines, and surfaces. The table is an example of disconnected discrete elements and therefore has the topology of discrete points. The timeseries is  sampled from a 1d continuous temperature measurement, and therefore has a topology of a 1D space. The map in \autoref{fig:related-work:continuity:ktypes} shows a 2D continuous surface. As illustrated in the figure, it is not uncommon for the same dataset to have multiple topologies. For the common topologies of discrete points, lines, and surfaces, Bertin recommend that they be represented visually by point, line, and area type visual elements respectively \cite{bertinSemiologyGraphicsDiagrams2011a}, and this expectation has been generalized in the development of visual algoriithms.

Visual algorithms assume the continuity of their data, as described in taxonomies of visualization algorithms Chi and by Troy and M\"{o}ller \cite{chiTaxonomyVisualizationTechniques2000,toryRethinkingVisualizationHighlevel2004} and try to preserve it, but many data structures do not provide a method for encoding the topology of the data they store. For example, a \texttt{line} algorithm often take as input a list of (x,y) coordinates. The \texttt{line} algorithm assumes that its input is continuous and therefore displays the data as if it was continuous, regardless of the actual data topology. The ability to encode the continuity expectations in the visual algorithm and the data container would provide a way of verifying that the assumptions match and therefore that the components are preserving the topological structure.

@@

\section{Formal Properties of Data \& Graphics}
\label{sec:atct}
In this section, we propose a mathematical abstraction of a data container and graphic prerendered output. This mathematical abstraction has a robust language for expressing topology and fields; expresses how to verify that data continuity is preserved on subset, distributed, and streaming data; and formalizes the expectation of a correspondence between data and visual elements.

\subsection{Uniform Abstract Data Representation}
\label{sec:atct:fiber-bundles}
To bridge the need for flexibility and cohesiveness, we model data using a mathematical representation of data that can encode topological properties, field types, and data values in a way that is both uniform and continuity and dimension independent way. This mathematical representation is called a fiber bundle and its use as an abstract data representation in visualization was first proposed by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}. A fiber bundle $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and continuous surjective map $\pi: \dtotalc \rightarrow \dbasec$\cite{FiberBundle2020}.

@@

The \textcolor{base}{base space} is a topological space \dbasec\ with points $\dbasepoint \in \dbase$ and topology $\mathcal{T}_{\dbase}$; $\mathcal{T}_\dbasepoint$ is a set of opensets $ \dbasepoint \in \openset \subseteqq \dbase$ \cite{munkresElementsAlgebraicTopology1984} that cover the base space \dbasec.

@@

 The pushforward $\vindexpushc$ transports sheaves of sections on $\opensetgc$ over $\openset$
 \begin{equation}  
  \vindexpushc\gsectionc: \opensetc \rightarrow \vindexpushc \gtotalc\restriction_{\opensetc} \in \cgamma{\opensetc}{\vindexpushc\gtotalc\restriction_{\opensetc}} 
\end{equation}
such that it provides a way to look up which graphic  corresponds with a data index
\begin{equation}
  \vindexpushc\gsectionc(\dbasepointc) = \gsectionc\restriction_{\vindexprec(\dbasepointc)} = \{\gsectionc(\gbasepointc)\;for\;all\; \gbasepointc\in \vindexprec(\dbasepointc)\}
\end{equation}
\note{$\xi_*rho(k)(s)=rho(s) for s in \xi^{-1}(k)$
 and move {} into narrative}
Therefore, the continuous map $\vindex$ and transport functors $\vindexpull, \vindexpush$ allow us to express the correspondence between graphic section and data section.

>>>>

\section{Related Work}
We build on the extensive work codifying the structure of visualization data to develop a more generalizable method for expressing the structure that visualizations are expected to preserve. Broadly, previous work describes structure as a combination of the topological properties of the data, the mathematical structure of the data fields, and equivalent changes to data and graphics. We propose that a unified abstraction model can guide developers in building structure aware composable, reusable software library components with more consistent interfaces.

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{
  A a topological representation of the data (in purple) is a simplified representation of a view of how the records in a dataset are connected. The table can be viewed as a set of independent records; therefore it has a 0D continuity represented by the disconnected dots. These weather records are also samples of continuous measurements at each location, which can be modeled as the time series measurements being samples on a 1d continuous interval and the geospatial station measurements as sparse samples on a 2D continuous space. 
  \label{fig:related-work:continuity:ktypes}}
\end{figure}

There is an assumption in visualization that the connectivity of the elements in a dataset, which is roughly what topology encodes\cite{wilkinsonGrammarGraphics2005}, are preserved in the visual representation. Bertin gives the examples of discrete data values should be represented by discrete entities such as points, quantitive continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011a}. In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and it is expected  that each rows will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved. Topological continuity allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that continuity in the visual representation beyond points, lines, areas, and other case by case continuities.

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi and by Troy and M\"{o}ller \cite{chiTaxonomyVisualizationTechniques2000,toryRethinkingVisualizationHighlevel2004}. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Since topological continuity is generalizable, we propose the construction of topology types such that data can explicitly state its topology type and visual algorithms can check that this type matches their assumptions.

@@

\section{Formal Properties of Data \& Graphics}
\label{sec:atct}
In this section, we propose a mathematical abstraction of the data input and graphic prerendered output. This mathematical abstraction has a robust language for expressing topology and fields; expresses how to verify that data continuity is preserved on subset, distributed, and streaming data representations; and formalizes the expectation of a correspondence between data and visual elements.

\subsection{Uniform Abstract Data Representation}
\label{sec:atct:fiber-bundles}
We model data using a mathematical representation of data that can encode topological properties, field types, and data values in a uniform. This mathematical representation is called a fiber bundle and its use as an abstract data representation in visualization was first proposed by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}. A fiber bundle $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and continuous surjective map $\pi: \dtotalc \rightarrow \dbasec$\cite{FiberBundle2020}.

@@

The \textcolor{base}{base space} is a topological space \dbasec\ with points $\dbasepoint \in \dbase$ and topology $\mathcal{T}_{\dbase}$. For our purposes, $\dbase$ is a set of points in a mathematical space-such as a metric space, manifold, or euclidean space-that encodes the underlying connectivity of the dataset. For example, $\dbase$ could be the discrete points, interval, or plane in \autoref{fig:related-work:continuity:ktypes}. On this set \dbase, we can define subsets of connected elements, called opensets \openset. An openset \openset\ is a generalization of an open interval, meaning an openset contains the points inside a bounded region on X and excludes the points on the boundary. A topology $\mathcal{T}_{\dbase}$ defined on \dbase\ is a collection of opensets such that the empty set and \dbase are in $\mathcal{T}_{\dbase}$,and every union and intersection of opensets in $\mathcal{T}_{\dbase}$ are also in $\mathcal{T}_{\dbase}$ \cite{TopologyVsTopology,bradleyTopologyCategoricalApproach2020}. The topology is encoding the notion that the connectivity is expected to be preserved regardless of how small an openset or large a union of opensets.

@@

 The pushforward $\vindexpushc$ transports sheaves of sections on $\opensetgc$ over $\openset$
 \begin{equation}  
  \vindexpushc\gsectionc: \opensetc \rightarrow \vindexpushc \gtotalc\restriction_{\opensetc} \in \cgamma{\opensetc}{\vindexpushc\gtotalc\restriction_{\opensetc}} 
\end{equation}
such that it provides a way to look up which graphic  corresponds with a data index
\begin{equation}
  \vindexpushc\gsectionc(\dbasepointc) = \gsectionc\restriction_{\vindexprec(\dbasepointc)} = \{\gsectionc(\gbasepointc)\;for\;all\; \gbasepointc\in \vindexprec(\dbasepointc)\}
\end{equation}
\note{\ci_*rho(k)(s)=rho(s) for s in \xi^{-1}(k)
} and move {} into narrative
Therefore, the continuous map $\vindex$ and transport functors $\vindexpull, \vindexpush$ allow us to express the correspondence between graphic section and data section.
